---
output:
  rmarkdown::html_document:
    highlight: pygments
    toc: true
    toc_depth: 3
    fig_width: 5
vignette: >
  %\VignetteIndexEntry{Bioc2020RCWL-1}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding[utf8]{inputenc}
---

```{r}
knitr::opts_chunk$set(message = FALSE)
```

# Connecting Bioconductor to other bioinformatics tools using `Rcwl`

## Instructor(s) name(s) and contact information

* [Qian Liu](https://github.com/liubuntu) (Qian.Liu@roswellpark.org)
* [Qiang Hu](https://github.com/hubentu) (Qiang.Hu@roswellpark.org)

## Workshop Description

This workshop introduces the Bioconductor toolchain for usage and
development of reproducible bioinformatics pipelines using packages of
`Rcwl` and `RcwlPipelines`. The Common Workflow Language (CWL) is an
open standard for the development of data analysis workflows that is
portable and scalable across different tools and working
environments. `Rcwl` provides a simple way to wrap command line tools
and build CWL data analysis pipelines programmatically within R. It
increases the ease of use, development, and maintenance of CWL
pipelines, and offers a higher performance by intuitively supporting
parallel work with high performance computing (HPC). More than a
hundred of pre-built and tested CWL tool/pipeline recipes are included
in `RcwlPipelines` package. These tools and pipelines are highly
modularized for easy customization of complex bioinformatics
analysis. Here we included a scRNAseq preprocessing pipeline that uses
`STARsolo` for alignment and quantification, and `DropletUtils` for
filtering raw gene-barcode matrix and removing empty droplets. This
pipeline demonstrates the typical use case of our packages. More
details for usage and examples are available on Rcwl website:
https://hubentu.github.io/Rcwl/.


## Pre-requisites

- Basic knowledge of using R and Bioconductor packages
- Basic familiarity with running command-line tools
- No prior experience with CWL is necessary!

## Workshop Participation

Participants will be able to try out all of the functionality
described. Active user participation throughout the event is highly
encouraged including but not limited to lecture material, hands-on
sections and final discussion.

Some basic ideas about how CWL works.
* https://www.commonwl.org/user_guide/

## _R_ / _Bioconductor_ packages used

* [`Rcwl`](https://bioconductor.org/packages/Rcwl/)
* [`RcwlPipelines`](https://bioconductor.org/packages/RcwlPipelines/)

System dependencies
* [cwltool](https://github.com/common-workflow-language/cwltool)
* [docker](https://docs.docker.com/get-docker/)

## Time outline

| Activity                                                      | Time |
|---------------------------------------------------------------|------|
| Overview of bioinformatics pipelines and CWL (ppt)            | 10 min |
| Use core functions to update, search and load tools/pipelines | 5 min  |
| Use single cell indexing tool `STARindex`                     | 5 min |
| Use single cell alignment tool `STARsolo`                     | 5 min  |
| Use single cell filtering tool `DropletUtils`                 | 5 min  |
| Use single cell preprocessing pipeline `STARsoloDropletUtils` | 5 min  |
| Wrap command line tools using `Rcwl`                          | 0m (developers only) |
| Customize your pipelines using `Rcwl`                         | 0m (developers only) |

## Workshop goals and objectives

### Learning goals

* Basic knowledge of Common Workflow Language (CWL)
* Knowledge of R/Bioconductor interface of CWL
* Usage of the pre-built bioinformatics tools and pipelines in _R_
* Understand how to wrap command line tools with `Rcwl` (developers only)
* Understand how to build bioinformatics pipelines with `Rcwl` (developers only)

### Learning objectives

* Run the scRNA-seq indexing tool `STARindex` in _R_
* Run the scRNA-seq alignment tool `STARsolo` in _R_
* Run the scRNA-seq QC tool `DropletUtils` in _R_
* Run the scRNA-seq preprocessing pipeline which combines alignment and QC steps in _R_
* Create a basic echo tool using `Rcwl` (developers only)
* Build a basic DNAseq pipeline using `RCWL` (developers only) 

## Introduction to RCWL & RcwlPipelines

Please see the details in the slides. 

## single cell RNA sequencing data 

### Data resources

We have prepared all the files needed in this workshop. By initiating
the docker image for this workshop, you should already have it
available in the filepath (`path`) below. We have also mounted your
local working directory to this rstudio session (`outpath`) to store
the intermediate or result files from running the examples.

```{r}
path <- "~/inst/testdata"  ## source data
outpath <- "~/outdir"      ## output files
dir.create("outdir", showWarnings = FALSE)
```

The single cell data resource we are using today is the 1k PBMCs from
10x genomics, which consists of 1000 Peripheral blood mononuclear cells
(PBMCs) extracted from a healthy donor, where PBMCs are primary cells
with relatively small amounts of RNA (~1pg RNA/cell).

The source material consists of 6 FASTQ files split into two
sequencing lanes L001 and L002, each with three reads of R1
(barcodes), R2 (cDNA sequences), I1 (illumina lane infold).  10x
Genomics has its own processing pipeline, `Cell Ranger` to process the
scRNA-seq outputs it produces (and requires many configurations to run
and is significantly slower than other mappers), which requires all
three files to perform the demultiplexing and quantification, but
`STARsolo` does not require the I1 lane file to perform the
analysis. The source files can be found
[here](https://zenodo.org/record/3457880#.XxLjOvhKjvU).

For this tutorial, we will use datasets sub-sampled from the source
files to contain only 15 cells instead of 1000. We have also further
curated the data to only include reads on chromosome 21, so that the
real execution of our CWL tools/pipelines in _R_ can be done within
1~2 minutes for each step. 

```
subset15_chr21_pbmc_1k_v3_S1_L001_R1_001.fastq.gz
subset15_chr21_pbmc_1k_v3_S1_L001_R2_001.fastq.gz
subset15_chr21_pbmc_1k_v3_S1_L002_R1_001.fastq.gz
subset15_chr21_pbmc_1k_v3_S1_L002_R2_001.fastq.gz
```

For the mapping, we need a “whitelist” of known cell barcodes. Here we
used the 15 barcodes that we already subsetted. The original 737,000
barcodes can be freely extracted from the [Cell Ranger
pipeline](https://kb.10xgenomics.com/hc/en-us/articles/115004506263-What-is-a-barcode-whitelist).

```
subset15_demo_barcode.txt
```

The barcodes in the R1 FASTQ data are checked against these known cell
barcodes in order to assign a specific read to a specific known
cell. The barcodes are designed in such a manner that there is
virtually no chance that they will align to a place in the reference
genome. 

In this tutorial, we will be using the hg19 (GRCh37) version of the
human genome, and will therefore also need to use a hg19 GTF file to
annotate our reads (on chromosome 21 only for this demo).

```
chr21.fa
Homo_sapiens.GRCh37.75.21.gtf
```

### `Rcwl` version of scRNA-seq preprocessing steps

In this tutorial, we are basically trying to reproduce the [GALAXY
pipeline](https://galaxyproject.github.io/training-material/topics/transcriptomics/tutorials/scrna-preprocessing-tenx/tutorial.html#producing-a-count-matrix-from-fastq)
for scRNA-seq data pre-processing. We will use `STARsolo` to produce a
count matrix from FASTQ, and `DropletUtils` to produce a high-quality
count matrix with feature/cell annotation files saved in an R object
of `SingleCellExperiment`. Before these 2 steps, we have also added a
one-time indexing step in our workshop.

With `Rcwl`, we can easily write cwl tools and pipelines
programmatically in _R_ (the other vignette called
"Bioc2020RCWL4devel" covers more of this topic for developers). We
have pre-built and tested more than a hundred of tools/pipelines which
can be loaded using `RcwlPipelines` (tools and pipelines developing
scripts can be found here:
https://github.com/hubentu/RcwlRecipes/tree/master/Rcwl). 

For usage of the existing tools/pipelines, Three major steps are
needed: 1) Search and load the tools/pipelines, 2) assign values for
each of the defined parameters, 3) execute the tools/pipelines. All
steps are done programmatically in _R_, and then we can get the
results ready in the user-specified directory.

#### Loading the tools

Here we show the usage of 3 core functions: `cwlUpdate`, `cwlSearch`
and `cwlInstall` for updating, searching, and loading the needed tools
or pipelines in _R_.

- `cwlUpdate`

The `cwlUpdate` syncs the current `Rcwl` recipes and return a
`BiocFileCache` object which contains the most updated Rcwl recipes.

```{r}
library(Rcwl)
library(RcwlPipelines)
```

```{r}
atls <- cwlUpdate()  ## sync the tools/pipelines
atls
```

the `bfcinfo()` function returns a BiocFileCache tibble containing all
information about each available tool or pipeline. Currently, we have
integrated 103 command line tools and 26 pipelines.

```{r}
bfcinfo(atls)
table(bfcinfo(atls)$Type)
```

- `cwlSearch`

We can use (multiple) keywords to search for any specific tools/pipelines of
interest, which internally search the keywords against the
BiocFileCache tibble in columns of "rname", "rpath", "fpath",
"Command" and "Containers". 

```{r}
tls <- cwlSearch(c("STAR", "index"))
data.frame(tls)
```

- `cwlInstall`

The last core function is `cwlInstall` which sources the `Rcwl` script
for the interested tool/pipeline into the _R_ working environment.

```{r}
cwlInstall(tls$rname[2])  ## "tl_STARindex"
cwlInstall(tls$fpath[2])  ## equivalent to the above. 
STARindex
```

Now the Rcwl version of `STARindex` is successfully loaded into _R_
and ready to use. 

#### Indexing

Before the alignment and QC in the aboved mentioned GALAXY pipeline,
we will add an indexing step using the tool `STARindex`. The command
line for indexing looks like this:

```
$ STAR --runMode genomeGenerate --runThreadN 4 --genomeDir STARindex 
--genomeFastaFiles chr21.fa --sjdbGTFfile Homo_sapiens.GRCh37.75.21.gtf
```

Using the `Rcwl` version of `STARindex`, we can equivalently do the
same indexing within _R_ which was internally passed as a cwl
script. What we need to do is assign values for the input parameters,
and execute the cwl script using `runCWL` function.

```{r} 
cwlVersion(STARindex)
inputs(STARindex)
outputs(STARindex)
requirements(STARindex)
```

Now let's assign values for these parameters. 

```{r, eval=FALSE}
inputs(STARindex)
STARindex$genomeFastaFiles <- file.path(path, "chr21.fa")
STARindex$sjdbGTFfile <- file.path(path, "Homo_sapiens.GRCh37.75.21.gtf")
STARindex  ## values (filepatt) are added
writeCWL(STARindex, prefix = file.path(outpath, "STARindex"))  ## 2 files .cwl/.yml 
```

The `runCWL(docker = )` takes 3 values: 
- TRUE: on the HPC, without the permission to pre-install all command line tools needed.   
- FALSE: on your local computer, with all command line tools pre-installed. 
- "singularity": On the HPC, no pre-installed tools, and without support of docker. 

```{r, eval=FALSE}
system.time(
    runCWL(cwl = STARindex, outdir = file.path(outpath, "STARindex_output"),
           docker = "singularity", showLog = TRUE)
    ) ## elapsed: 74.795S
dir(file.path(outpath, "STARindex_output"), recursive = TRUE)
```

Now that the output files are generated in the folder of
"STARindex_output/STARindex" under `outpath`. These output files are
ready to pass as input to the next tool for single cell read
alignment.

#### Alignment

To align single cell reads using `STAR`, it can be done directly in
command line:

```
$ STAR --soloType CB_UMI_Simple --genomeDir STARindex_output/STARindex --soloCBwhitelist subset15_demo_barcode.txt --soloUMIlen 12 --readFilesIn subset15_chr21_pbmc_1k_v3_S1_L001_R2_001.fastq.gz,subset15_chr21_pbmc_1k_v3_S1_L002_R2_001.fastq.gz subset15_chr21_pbmc_1k_v3_S1_L001_R1_001.fastq.gz,subset15_chr21_pbmc_1k_v3_S1_L002_R1_001.fastq.gz --soloUMIfiltering MultiGeneUMI --soloCBmatchWLtype 1MM_multi_pseudocounts --readFilesCommand gzcat --runThreadN 1
```

Now we can follow the previous example to load our `Rcwl` version of
`STARsolo` and run the command in _R_.

```{r}
cwlInstall("tl_STARsolo")  ## "tl_STARsolo"
STARsolo
inputs(STARsolo)
```
```{r, eval=FALSE}
cdna.fastq <- file.path(path, list.files(path, pattern = "_R2_"))
cb.fastq <- file.path(path, list.files(path, pattern = "_R1_"))
cblist <- file.path(path, "subset15_demo_barcode.txt")
genomeDir <- file.path(outpath, "STARindex_output/STARindex")

STARsolo$readFilesIn_cdna <- cdna.fastq
STARsolo$readFilesIn_cb <- cb.fastq
STARsolo$whiteList <- cblist
STARsolo$genomeDir <- genomeDir
STARsolo
writeCWL(STARsolo, prefix = file.path(outpath, "STARsolo"))
```

```{r, eval=FALSE}
runCWL(STARsolo, outdir = file.path(outpath, "STARsolo_output"),
       docker = "singularity", showLog = TRUE)  ## approximately 2 mins
dir(file.path(outpath, "STARsolo_output"), recursive = TRUE)
```     

Now we have all files generated in the "STARsolo_output" folder under
`outpath`, which can be passed into the next tool.

#### `DropletUtils`

To get a high quality count matrix we must apply the `DropletUtils`
tool, which will produce a filtered dataset that is more
representative of the Cell Ranger pipeline.

Since CWL itself doesn't support the integration of _R_ packages or
_R_ function, and lacks interoperability of very widely used _R_ tools
for downstream data analysis. This is a unique feature for `Rcwl`,
where we can easily connect the upstream data preprocessing steps
(previously based on command line tools) and the downstream data
analysis steps (heavily done in _R/Bioconductor_).

The idea here is to put anything you want to do into an _R_ function,
with specified arguments for input and output files, then it's ready
to be wrapped as an `Rcwl` tools for execution. 

For example, in using the package `DropletUtils`
(https://github.com/hubentu/RcwlRecipes/blob/master/Rcwl/tl_DropletUtils.R),
we mainly did 3 things: 1) use the `read10xCounts` function to read
the output files from the alignment step and convert into a
`SingleCellExperiment` object. 2) Calculate the barcode ranks and
plotting. 3) calcualte the empty droplets and plotting.

```{r}       
cwlInstall("tl_DropletUtils")
inputs(DropletUtils)
```

```{r, eval=FALSE}
DropletUtils$dirname <- file.path(outpath, "STARsolo_output/Solo.out")
DropletUtils$lower <- 100
DropletUtils$df <- 5
```

```{r, eval=FALSE}
runCWL(DropletUtils, outdir = file.path(outpath, "dropletUtils_output"), showLog = TRUE)
dir(file.path(outpath, "dropletUtils_output"))
```

Now that we get 2 output files:

- The pdf file with 2 diagnostic figures: Barcode ranks, and empty droplets. Details for interpretation of each diagnostic figure please refer to the [DropletUtils vignette](https://bioconductor.org/packages/release/bioc/vignettes/DropletUtils/inst/doc/DropletUtils.html)  
- The `SingleCellExperiment` object which has filtered out unqualified cells.  

```{r, eval=FALSE}
#> Biobase::openPDF(file.path(outpath, "dropletUtils_output/diagnostics.pdf"))
sce <- readRDS(file.path(outpath, "dropletUtils_output/sce_filtered.rds"))
sce
```

### scRNA-seq preprocessing pipeline

Alternatively and more easily, we can connect these tools and make a
pipeline, which is already available in `RcwlPipelines` called
`pl_STARsoloDropletUtils`. It has integrated the `STARsolo` and
`DropletUtils` for a streamlined pre-processing analysis within
_R_. These pipelines are ready for customization for your own
research.

In a pipeline, we only need to assign input values for the whole pipeline, not individual tools involved. The input and output between each step are pre-defined in the pipeline to ensure a smooth passing.  


```{r}
cwlInstall("pl_STARsoloDropletUtils")
plotCWL(STARsoloDropletUtils)
```

```{r, eval=FALSE}
inputs(STARsoloDropletUtils)
STARsoloDropletUtils$fastq_cdna <- cdna.fastq
STARsoloDropletUtils$fastq_cb <- cb.fastq
STARsoloDropletUtils$genomeDir <- file.path(outpath, "STARindex_output/STARindex")
STARsoloDropletUtils$whiteList <- cblist
STARsoloDropletUtils$runThreadN <- 1
```

```{r, eval=FALSE}
runCWL(STARsoloDropletUtils, outdir = file.path(outpath, "scpipeline_output"),
       docker = "singularity", showLog = TRUE)
```

The overall output of the pipeline was pre-defined to glob the
important files from separate steps.

```{r, eval=FALSE}
dir(file.path(outpath, "scpipeline_output"), recursive = TRUE)
outputs(STARsoloDropletUtils)
```

### Summary and some additional comments

There are some other cool functionalities that we haven't included in
this workshop. For example, the `runCWLBatch()` function is designed
for execution of CWL pipelines in high performance computing with
support of different job submitting systems, such as slurm, SGE,
etcs. The `cwlShiny` opens a user-friendly shiny interface for any
`Rcwl` tools.

As a summary, this workshop we have introduced the usage of two
packages: `Rcwl` and `RcwlPipelines` in constructing and executing the
CWL tools/pipelines with _R_ for the previously command line tools as
well as customized _R_ functions. The pre-built tools and pipelines
are highly modularized and optimized for easy customization for
specific data analysis needs. These packages are under active
development, and we welcome any question for the functionalities,
feature requests (support site, email), and issue reports
(https://github.com/hubentu/Rcwl/issues,
https://github.com/hubentu/RcwlPipelines/issues). 
importantly, pull requests for adding your sp

Most importantly, we are trying to make this project as a community
effort for developing and sharing of specific sets of tools and
pipelines in their bioinformatics domains. We look forward to any
collaborations in developing the pipelines and please feel free to
make your pull requests for your recipes
[here](https://github.com/hubentu/RcwlRecipes). 

### Acknowledgement

This work was supported by Clinical and Translational Science Award
[UL1TR001412 or KL2TR001413] to the University at Buffalo from the NIH
National Center for Advancing Translational Sciences.

